Real-time Acoustic Ray Range Estimation for Underwater Vehicle Navigation

### 2021-08-26 doc/main.tex
draft shared with Henrik + Oscar

\title[JASA/draft]{Real-time Acoustic Ray Range Estimation for Underwater Vehicle Navigation}
\author{EeShan C. Bhatt}
\email{ebhatt@whoi.edu}
\affiliation{MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge and Woods Hole, MA, USA}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
\author{Oscar Viquez}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
\author{Bradli Howard}
\affiliation{MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge and Woods Hole, MA, USA}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
\author{Henrik Schmidt}
\email{henrik@mit.edu}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}

@@

\begin{abstract}
Almost all underwater navigation paradigms rely on the conversion of recorded travel time into range to compute a trilateration solution. For real-time navigation, this conversion is treated deterministically. For positioning in post-processing, more computationally and/or labor intensive modeling methods are often employed to reduce the uncertainty driven by multipath arrivals. For an autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea, the total under-ice environment and complex acoustic propagation via the Beaufort Lens necessitated the real-time integration of model-based data processing. An embedded, stochastic, ray-based prediction of the horizontal group velocity achieved a ranging mean absolute error of 11 m compared to GPS-aided beacon-to-beacon communication events. In post-processing, a multipath adjustment achieved ranging performance that rivals GPS. To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based model to constrain ranging error in underwater navigation.
\end{abstract}

@@

Much work has been done in underwater vehicle navigation, which relies on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and/or acoustic baseline navigation systems \citep{paull_auv_2014}.
In particular, the long baseline (LBL) navigation system is the most similar in scale and style to GPS, and most appropriate for mitigating drift in underwater vehicle localization without overburdening computation on the vehicle \citep{van_uffelen_global_2021}.

@@

Positioning is the ability to accurately determine one's location and is generally done via trilateration.
Real-time navigation efforts assume a locally homogeneous sound speed profile (SSP) that enables a linear, deterministic scaling between travel time and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
Short range (less than ten kilometer) navigation efforts accurately rely on a nominal sound speed value and achieve minimal navigation error \cite{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
For mesoscale (tens to hundreds of kilometers) navigation or localization efforts, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed is used.

Post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), without computational or time constraints, can indulge more complex acoustic methods and/or ocean representations.
There are many different approaches, from accounting for a depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}, or isolating a specific ray path to find a representative group velocity \citep{mikhalevsky_deep_2020}.
The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.

@@

However, the key assumptions for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash did not hold during a vehicle deployment in ICEX16, in the Beaufort Sea, in March 2016.

However, the new Arctic ... [Beaufort Lens]

The literature in underwater acoustic navigation and positioning is either real-time or physics-based.
AUV operations in the Beaufort Sea necessitate a real-time and physics-based method to overcome the unique challenges of total ice cover, multiple ice floes, and the Beaufort Lens.

The paper is organized as follows: Section~\ref{sec:2} presents the physical
experimental setup and data processing simplifications; Section~\ref{sec:3} presents key results for group velocity predictions and range estimations; Section~\ref{sec:4} closes with a discussion.

% =========================================================================== %
% =========================================================================== %
\section{\label{sec:2} Methods}

The operational solution for ICEX20 relied on an Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
The ICNN was initially developed via numerous virtual experiments, to push the boundaries of algorithms and interfaces between different hardware components.
The simulation approach is a necessary testbed for robustness and produces better results than post-processing previous field data, as that restricts mission configurations to the data taken, which can hamper troubleshooting in the field.
The simulation capabilities are largely physics-driven with a modular system of systems approach: an environmental simulator with sub-components for the ocean, Arctic ice, and ocean acoustics; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software and hardware-in-the-loop version \citep{schneider_netsim_2018}.
The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
Both simulation capabilities are integral to mission success.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=\reprintcolumnwidth]{figs/ICNN_Overview}
	\label{fig:icnnOverview}
	\caption{CAPTION}
\end{figure}

@@

Vehicle missions in any underwater environment, but especially acoustically complex ones, must balance competing uses of the acoustic channel.
This network uses a single synchronized digital communication packet to provide both tracking and data to the operator.
The ICNN operational paradigm is as follows:
\begin{enumerate}
\item The AUV, running an ice-tracking DVL and an onboard hydrodynamic model, broadcasts its perceived location on a scheduled, time-synchronized message via WHOI Micro-Modem
\item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{porter_bellhop_2011}
\item The topside computer calculates a new position by trilaterating the range estimates
\item The position differential is broadcast to the vehicle to update its navigation solution
\end{enumerate}

This paper specifically addresses the third step\textemdash the topside computer's conversion of travel time into range, which underpins the trilateration in the ICNN, and is an often overlooked element of any underwater acoustic navigation system.
The results accordingly focus on validating acoustic range estimates from communication events between GPS-tracked beacons.
This ``modem experiment'' was conducted while the vehicle was still being prepped to go under the ice.

@@

Because the navigation solution on the vehicle, during a mission, can only be evaluated on the basis of the error estimates sent, a sister experiment for validating the real-time ranging approach was implemented.
The ice buoy modems were run as ``virtual vehicles'' at a given depth and would transmit to all other ice buoys.
The internal sound speed estimate was then toggled using a handful of weights applied to a basis function representation of the sound speed.
This trial was configured for as many ice buoy modems as possible, with each experiment running for roughly an hour between the two sound speed modalities.
There are three important engineering choices in the ICNN network to contextualize this dataset:
\begin{enumerate}
\item Each modem buoy only has one transmit depth layer (30 or 90 m)
\item The transmit and receive layers are independent
\item At any given time, the receive depth layers are the same across all four ice buoys
\end{enumerate}

The design enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
This adaptivity was not used during the modem experiment but was instead toggled manually.
Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive layer, and sound speed combinations, as this time was also used to verify and resolve any hardware issues in the modems themselves.

Importantly, only one modem at a time could run the vehicle behavior to trigger group velocity calculations; other modems that were transmitting simultaneously recorded events as modems, not as vehicles.

The dataset introduced in this thesis has parallel physical and virtual layer; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.

\subsubsection{Physical layer}
The physical layer, i.e. the acoustic link through the real ocean, describes the beacon to beacon communication events through the water column; it covers all source depths (20, 30, and 90 m) and both receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS. Events are sorted by source depth and sound speed profile, given that these are the dominating factor for the location and span of the shadow zone compared to any temporal or spatial scales of variability.

\begin{figure}[h!]
  \centering
  \includegraphics[width=.50\textwidth]{figs/modem-chart.pdf} \hfill
  \includegraphics[width=.48\textwidth]{figs/zs20-birdseye.pdf} \\
  \vspace{1em}
  \includegraphics[width=.48\textwidth]{figs/zs30-birdseye.pdf} \hfill
  \includegraphics[width=.48\textwidth]{figs/zs90-birdseye.pdf} \\
  \caption{An overview of the modem experiment by source and receiver depth and position. Subfigure (a) shows a chart of the successful events between source and receiver depth. The black column on the left, $tx$, shows the source depth, $z_s$. The column on the right, $rx$, shows the receivers with the amount of good contacts. The orientation of the triangles\textemdash sideways, upwards, and downwards\textemdash corresponds to depths of 20, 30, and 90 m. Subfigures (b-d) correspond to receiver depths of 20, 30, and 90 m, respectively. Camp Seadragon, demarcated by the purple triangle, is situated just north of the continental shelf.}
  \label{fig:overview}
  \end{figure}

@@

\subsubsection{Virtual layer}
The virtual layer describes the BELLHOP simulations for each communication event in the physical layer, where a handful of weights applied to a basis function representation can reconstitute a number of realistic candidate SSPs \citep{leblanc_underwater_1980,lin_merging_2010}.
The SSP is the only configurable input to the group velocity prediction.

During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a CTD, i.e. an optimal set of weights chosen by a human decision maker via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data.
This order corresponds to increasing ducted conditions of the Beaufort Lens.

Fig. \ref{fig:raytrace-zs30} shows eigenrays in a range independent coordinate system for all three sound speed environments for a source depth of 30 m.
It is important to note that BELLHOP produces an abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
The important pattern is the increased complexity in multipath via surface interactions as ducted conditions increase.

@@

The variation of horizontal group velocity from any source-receiver pair is the fundamental challenge to implement GPS-like navigation in an LBL navigation paradigm, especially in an acoustically complex propagation environment.
The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and range modems, supported by clock synchronization and a pre-determined scheduling of acoustic events.
For the Beaufort Lens, the strong multipath effects make it virtually impossible to deterministically predict the modem triggering time.

\subsubsection{Minimal bounce criteria}
Instead, for each individual node $i$, an embedded stochastic tracking framework is used to provide a running estimate of the horizontal group velocity $u_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the naive group velocity, i.e. the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.

@@

The initial call to BELLHOP is over a sparse, dynamically created depth and range grid, of which the plane wave approximation is used to bring the ray to the range and depth posited by the onboard tracking solution.
Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current group velocity $u$ from a power weighted average of the ray travel times,
\begin{equation}
u = \frac{\hat{r} \sum_{n=1}^{N_{0}} a_{n}^{2}}{\sum_{n=1}^{N_{0}} dt_{n}a_{n}^{2}} ~, 
\end{equation}
and the associated weighted standard deviation,
\begin{equation}
\sigma_{u} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
\end{equation}
If no direct paths exist, i.e. $N_{0}=0$, then the group velocity is computed using the same algorithm for the ray arrivals with one bounce, and so on.

@@

This stochastic method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than general post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
The method assumes direct path arrival, where multipath over the shallow receiver depth is trivial for the group velocity calculation; thus we call it the minimal bounce criteria (MBC).
The BELLHOP simulation that runs this calculation uses 3600 rays with launch angle fan of -60 to 60 degrees.
However, this high density of launch angles creates a greedy estimator of direct path eigenrays, and this simulation pipeline almost always guarantees a direct path found.
The MBC improves upon a deterministic estimate in that it considers the ray path, travel time, and amplitude to predict a realistic group velocity.

@@

As shown in the eigenray traces of Fig. \ref{fig:raytrace-zs30}, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
Often, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
The suggested algorithm, the nearest bounce criteria (NBC), is a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.
This metric was introduced in post-processing.

@@

This method selects a different group velocity based on the multipath arrival structure, as the expected arrival is not always the first arrival and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.

@@

\FloatBarrier
\section{\label{sec:4} Real-time psuedorange results}

The modem experiment generated 811 beacon to beacon communication events with their own real-time NBC group velocity predictions.
As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.
This section examines the absolute range error between the psuedorange and the GPS derived range with the caveat that the data collected is fairly asymmetrical across source depths and sound speed conditions.

@@

The minimal and nearest bounce algorithm is applied using the three sound speed inputs shown in Fig.~\ref{fig:raytrace-zs30}; the resulting predicted group velocities for a source depth of 30 m are shown in Fig~\ref{fig:gvel30}.
The takeaways from this source depth are representative of those from 20 and 90 m as well (see Supplementary Fig.~\ref{fig:gvelMore}).

@@

% =========================================================================== %
% =========================================================================== %
\section{\label{sec:4} Discussion}

>>>>

\title[JASA/draft]{Real-time Acoustic Ray Range Estimation for Underwater Vehicle Navigation}
\author{EeShan C. Bhatt}
\email{ebhatt@whoi.edu}
\affiliation{MIT-WHOI Joint Program in Oceanography/Applied Ocean Science \& Engineering, Cambridge and Woods Hole, MA, USA}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
\author{Oscar Viquez}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}
\author{Henrik Schmidt}
\email{henrik@mit.edu}
\affiliation{Department of Mechanical Engineering, Massachusetts Institute of Technology, Cambridge, MA}

@@

\begin{abstract}
Almost all underwater navigation paradigms rely on the conversion of recorded travel time into range to compute a trilateration solution.
For real-time navigation, this conversion is treated deterministically.
For positioning in post-processing, more computationally and/or labor intensive modeling methods are often employed to reduce the uncertainty driven by multipath arrivals.
An autonomous underwater vehicle deployment in March 2020 in the Beaufort Sea faced significant operational challenges with a total ice-covered environment and complex acoustic propagation via the Beaufort Lens.
We implement an embedded, stochastic, ray-based prediction of the horizontal group velocity and validate its performance with communication events between GPS-aided beacons, achieving a psuedorange mean absolute error of 11 m.
In post-processing, we introduce a multipath criterion and achieve ranging performance that rivals GPS.
To our knowledge, these results are the first field experiment to demonstrate a real-time, physics-based data processing to minimize ranging error in underwater navigation.
\end{abstract}

@@

Much progress has been made in underwater vehicle navigation, which relies on any combination of dead reckoning, hydrodynamic models, inertial navigation systems, doppler velocity logs, and/or acoustic baseline navigation systems \citep{paull_auv_2014}.
In particular, the long baseline (LBL) navigation system is the most similar in scale and style to GPS, and most appropriate for mitigating drift in underwater vehicle localization without overburdening computation on the vehicle \citep{van_uffelen_global_2021}.

@@

Positioning is the ability to accurately determine one's location and is generally done via trilateration.
Navigation assumes a degree of autonomy to move from a current position to a desired one.
The standard for real-time navigation efforts assume a locally homogeneous sound speed profile (SSP) that enables a linear, deterministic scaling between travel time and range \citep{eustice_recent_2006,eustice_experimental_2007,webster_preliminary_2009,webster_advances_2012}.
Short range (less than ten kilometers) navigation efforts achieve minimal navigation error with a nominal sound speed value \citep{eustice_experimental_2007,webster_preliminary_2009,kepper_mems_2017}.
For mesoscale (tens to hundreds of kilometers) navigation efforts, a deterministic \citep{graupe_preliminary_2019} or through-the-sensor \citep{webster_towards_2015} value for sound speed is used, but error remains in the hundreds of meters.

Post-processing localization efforts (re-navigation for a self-determinedly moving object or re-positioning for a passively moving one), without computational or time constraints, can indulge more complex acoustic methods and/or ocean representations.
There are many different approaches, from considering a temporally averaged mean depth dependent SSP \citep{van_uffelen_localization_2016}, including spatio-temporal variability in the sound speed profile to minimize error \citep{graupe_preliminary_2019, mikhalevsky_deep_2020}, comparing acoustic records with synthetic acoustic records computed through an ocean model \citep{wu_deep_2019}. 
The ``cold start'' algorithm developed by \citet{mikhalevsky_deep_2020} isolates a specific ray path to find a representative group velocity \citep{mikhalevsky_deep_2020} and re-positioned a floating hydrophone array with an error of 85 m and a standard deviation of 32 m relative to ground truth GPS data, but implementing this algorithm for navigation remains to be seen.
These post-processing efforts have generally been deemed too computationally intensive and/or labor intensive for real-time efforts, where automation of identifying time arrivals is further complicated by multipath.

@@

Underwater navigation research is broadly motivated by GPS-like navigation in GPS-denied conditions.
The Arctic, while remote, is the perfect place to test mature navigation technologies in real GPS-denied conditions.

Literature and historical reviews of underwater navigation approaches in under-ice conditions highlight the emerging technology learning curve for vehicle operations in the Arctic, the hostile conditions brought on by a semi-permanent moving ice cover, and the need for both absolute and relative navigation with little to no tolerance for error \citep{norgren_unmanned_2014,mcfarlane_autonomous_2015,barker_scientific_2020}.

One of the first successful AUV launches and recoveries involved equipping the vehicle with barbs which became entangled into a homing net underneath the ice \citep{bellingham_auv_1993}.
Since then, AUVs have been used for ocean bottom \citep{jakuba_long-baseline_2008,kunz_deep_2008}, under-ice \citep{kaminski_12_2010}, and hydrographic surveys \citep{kukulya_under-ice_2010,kukulya_development_2016,stevens_linking_2016,plueddemann_autonomous_2012}.
These efforts show a combination of short and long range positioning systems, where all acoustic navigation methods used a depth-averaged sound speed for linear scaling between recorded travel time and range.
In many cases robust quantitative navigation performance could not be evaluated due to small operational scale and/or moving ice floes.
But most importantly, all of these efforts experienced a nearly isovelocity or classical upward refracting Arctic acoustic environment.

The new Arctic\textemdash one with much less multi-year sea ice\textemdash is experiencing significant changes in the water column as well \citep{mackinnon_warm_2021}.
The Beaufort Lens is a highly dynamic, highly stratified water mass, keeping a neutrally buoyant layer of warm Pacific water at 50 to 70 m in depth.
Surrounded by cooler waters, it distinguishes a local maximum in the SSP, creating an upper and lower duct\textemdash the upper duct degrades signal coherence due to intensified ice interaction and the lower duct effectively traps sound for long range propagation \citep{poulsen_acoustic_2016}. 
Modeling output \citep{duda_long-range_2019,duda_effects_2021} and experimental observations \cite{kucukosmanoglu_observations_2021,badiey_azimuthal_2019,ballard_temporal_2020} indicate a high degree of spatio-temporal variability of the lower duct throughout the Chukchki and Beaufort Seas.

An experiment in 2014 exploited this lower duct to provide acoustic navigation and communication at mesoscale ranges in the Canada Basin.
\citet{freitag_long_2015} and \citet{webster_towards_2015} deployed multiple GPS-linked ice buoys with transducers that were capable of communications greater than 400 km with a range accuracy of 40 m when the sound speed was known.
The transducers were at at 100 m depth, well-positioned in the lower duct spanning 50 to 200 m.
The glider navigation accuracy, however, was severely degraded as acoustic packets were sent when the gliders were outside the sound channel.
When gliders were deployed again in 2017, their navigation error was on the order of half a kilometer, and a glider re-navigation using an acoustic arrival method and the average sound-speed profile from the gliders reduced navigation error by a factor of 4-5, depending on the dive \citep{graupe_preliminary_2019}.
Navigation error is referenced to intermittent GPS fixes, as all of these glider deployments occurred in the Marginal Ice Zone with little to no ice at the time.

\subsection{Ice Exercise 2020}

The Beaufort Lens is equally important for shorter ranges as well, where a shadow zone over vehicle operational ranges is created by a drastic change in the reliable acoustic path \citep{schmidt_acoustic_2016}.
Thus the key simplifying assumption for OWTT ranging\textemdash a locally homogeneous sound speed profile that enables a linear, deterministic scaling between travel time and range\textemdash becomes more incorrect due to increased multipath in the upper duct of the Beaufort Lens.

The results from this paper derive from data taken while deploying AUV Macrura, a modified Bluefin-21, during the Ice Exercise 2020 (ICEX20), in the Beaufort Sea, in March 2020.
The challenges anticipated during ICEX20\textemdash total ice cover, moving ice floes, and a variable Beaufort Lens\textemdash crystallized the necessity of environmental awareness for the navigation solution.

In this paper we present a working solution for underwater navigation that achieves GPS-like accuracy and precision.
We embed a stochastic prediction of a single group velocity to convert recorded travel time into range.
This prediction ingests an updatable snapshot of the SSP via acoustic message; over the course of ICEX20, this capability was used several times. 
Ultimately, integrating environmental realism enabled a successful emergency recovery of the AUV.
Due to a harddrive error, the AUV stalled underneath the ice surface and broadcasted its location.
A team was able to dig a hole through the 2 meter ice sheet and immediately see the vehicle.
Its successful recovery amidst total ice cover is shown in Fig. \ref{fig:vehicleRecovery}.
\begin{figure}[h!]
	\centering
	\includegraphics[width=\reprintcolumnwidth]{figs/vehicleRecovery.jpg}
	\caption{The successful emergency recovery of AUV Macrura, with the the auger used to drill through the ice sheet in the background. Photo courtesy of LT CDR D. Goodwin.}
	\label{fig:vehicleRecovery}
\end{figure}

The rest of this paper is organized as follows: Section~\ref{sec:methods} presents the physical
experimental setup and model-based data processing; 
section \ref{sec:realtime} quickly conveys real-time results and the limitations of the data collected; section~\ref{sec:postprocess} presents key results for group velocity predictions and range estimations; section~\ref{sec:discussion} closes with a discussion on importance, future work, and broader relevance.

% =========================================================================== %
% =========================================================================== %
\section{\label{sec:methods} Methods}

The acoustic communication dataset accrued for ICEX20 relied on the Integrated Communication and Navigation Network (ICNN) \citep{schneider_self-adapting_2020,randeni_construction_2020,randeni_high-resolution_2021}.
The ICNN was initially developed via numerous virtual experiments, to push the boundaries of algorithms and interfaces between different hardware components.
The simulation approach is a necessary testbed for robustness and produces better results than post-processing previous field data, as that restricts mission configurations to the data taken, which can hamper troubleshooting in the field.
The simulation capabilities are largely physics-driven with a modular system of systems approach: an environmental simulator with sub-components for the ocean, Arctic ice, and ocean acoustics; a vehicle simulator with sub-components for vehicle dynamics and navigation; a topside hardware simulator and acoustic communications simulator, both with a software and hardware-in-the-loop version \citep{schneider_netsim_2018}.
The virtual environment similarly emulates the interfaces between the real components to test the entire software pipeline.
Both simulation capabilities are integral to mission success.

@@

\begin{figure}[h!]
	\centering
	\includegraphics[width=\reprintcolumnwidth]{figs/ICNN_Overview}
	\caption{A cartoon overview of the Integrated Communication and Navigation Network (ICNN), which provides joint data and tracking information between AUV and a human decision maker at Topside.}
	\label{fig:icnnOverview}
\end{figure}

@@

Vehicle missions in any underwater environment, but especially acoustically complex ones, must balance competing uses of the acoustic channel.
This network uses a single synchronized digital communication packet to provide both tracking and data to the operator.
The ICNN operational paradigm is as follows:
\begin{enumerate}
\item The AUV, running an ice-tracking DVL and an onboard hydrodynamic model, broadcasts its perceived location on a scheduled, time-synchronized message via WHOI Micro-Modem
\item Four ice buoys, each outfitted with a WHOI Micro-Modem, receive messages from the AUV and send that information over freewave radio to a Topside computer
\item The topside computer converts travel times into range estimates using a stochastic embedded prediction of the horizontal group velocity via BELLHOP ray tracing code \citep{porter_bellhop_2011} coupled with a Virtual Ocean \citep{schneider_netsim_2018}
\item The topside computer calculates a new position by trilaterating the range estimates
\item The position differential, not the absolute position, is broadcast to the vehicle to update its navigation solution and be robust to latency and intermittency
\end{enumerate}

This paper specifically addresses the third step\textemdash the topside computer's conversion of travel time into range, which underpins the trilateration in the ICNN, and is an often overlooked element of any underwater acoustic navigation system.
The analysis accordingly focuses on validating acoustic range estimates from communication events between GPS-tracked beacons, not the AUV itself.
This allows the results to be decoupled from all the other moving parts in the ICNN.
This ``modem experiment'' was conducted while the vehicle was still being prepped to go under the ice.

@@

Because the navigation solution on the vehicle during a mission can only be evaluated on the basis of the error estimates sent, a sister experiment for validating the real-time ranging approach was implemented.
The ice buoy modems were run as ``virtual vehicles'' at a given depth and would transmit to all other ice buoys.
The internal sound speed estimate was then toggled using a handful of weights applied to a basis function representation of the sound speed.
This trial was configured for as many ice buoy modems as possible with the time allotted, with each experiment running for roughly an hour between the two sound speed modalities.
There are three important engineering choices in the ICNN network to contextualize this dataset:
\begin{enumerate}
\item Each modem buoy only has one transmit depth layer (30 or 90 m)
\item The transmit and receive layers are independent
\item At any given time, the receive depth layers are the same across all four ice buoys
\end{enumerate}

The design enables a self-adapting network to transmit and receive at the optimal depth to maintain contact with the AUV \citep{schneider_self-adapting_2020}.
This adaptivity was not used during the modem experiment but was instead toggled manually.
Given the complexity of the ICNN system, this experiment did not collect an exhaustive set of data across all buoy, source depth, receive layer, and sound speed combinations, as this time was also used to verify and resolve any hardware issues in the modems themselves.
Importantly, only one modem at a time could run the vehicle behavior to trigger group velocity calculations; other modems that were transmitting simultaneously recorded events as modems, not as vehicles.

A key construction in the ICNN are parallel physical and virtual layers; given the small operational scale, on the order of kilometers, and a singular SSP, both layers are projected into a range independent coordinate system for ease of comparison.

\subsubsection{Physical layer}
The physical layer describes the acoustic link between beacons in the water column.
It covers all source depths (20, 30, and 90 m) and both receive layers (30 and 90 m), where the surface expressions of the beacons are tracked by GPS.
Events are sorted by source depth and sound speed profile, given that these are the dominating factor for the location and span of the shadow zone compared to any temporal or spatial scales of variability.

\begin{figure}[h!]
  \centering
  \includegraphics[width=.51\textwidth]{figs/modem-chart.pdf} \hfill
  \includegraphics[width=.48\textwidth]{figs/zs20-birdseye.pdf} \\
  \vspace{1em}
  \includegraphics[width=.48\textwidth]{figs/zs30-birdseye.pdf} \hfill
  \includegraphics[width=.48\textwidth]{figs/zs90-birdseye.pdf} \\
  \caption{An overview of the modem experiment by source and receiver depth and position. Subfigure (a) shows a chart of the successful events between source and receiver depth. The black column on the left, $tx$, shows the source depth, $z_s$. The column on the right, $rx$, shows the receivers with the amount of good contacts. The orientation of the triangles\textemdash sideways, upwards, and downwards\textemdash corresponds to depths of 20, 30, and 90 m. Subfigures (b-d) correspond to receiver depths of 20, 30, and 90 m, respectively. Camp Seadragon, demarcated by the purple triangle, is situated just north of the continental shelf.}
  \label{fig:overview}
  \end{figure}

@@

\subsubsection{Virtual layer}
The virtual layer describes the BELLHOP simulations for each communication event in the physical layer.
The SSP used for the raytracing is driven by a handful of weights applied to a basis function representation
of the SSP variability to reconstitute a number of realistic candidate SSPs \citep{leblanc_underwater_1980,lin_merging_2010}.
The SSP is the only human configurable input to the group velocity prediction.

During ICEX20, the SSP was toggled between the historical``baseline'' and a ``weighted'' estimate based on a CTD, i.e. an optimal set of weights chosen by a human decision maker via an interactive Tactical Decision Aid framework \citep{bhatt_embedded_2021}.
For the post-processing analysis, we add the SSP from HYCOM for comparison, to mirror the information available on a submarine\textemdash modeled data, historical data, and \textit{in situ} data (personal conversation with LT B. Howard and LT CDR D. Goodwin).
This order corresponds to increasing ducted conditions of the Beaufort Lens.

Fig. \ref{fig:raytrace-zs30} shows eigenrays in a range independent coordinate system for all three sound speed environments, for a source depth of 30 m.
It is important to note that BELLHOP produces an abundance of eigenrays; the ones displayed here are chosen by travel time proximity to the recorded data.
With this filter, the chosen eigenrays show a greater amount of surface interactions as the SSP duct conditions increase.

@@

The variation of horizontal group velocity from any source-receiver pair is the fundamental challenge to implement GPS-like navigation in an LBL navigation paradigm, especially in an acoustically complex propagation environment.
The use of the acoustic modem network for tracking relies on the accurate estimates of travel times between the submerged platform and range modems, supported by clock synchronization and a pre-determined scheduling of acoustic events.
For the Beaufort Lens in particular, the strong multipath effects make it virtually impossible to deterministically predict the modem triggering time.

\subsubsection{Minimal bounce criteria (MBC)}
Instead, for each individual node $i$, an embedded stochastic tracking framework is used to provide a running estimate of the horizontal group velocity $u_{i,j}$ for the conversion from travel time to range from modem $j$, with the ultimate goal of matching the naive group velocity, i.e. the GPS-recorded distance between two nodes divided by the modem-recorded one way travel time between them.

@@

The initial call to BELLHOP is over a sparse local grid, of which the plane wave approximation is used to bring the ray to the range and depth posited by the onboard tracking solution.
For each grid point, BELLHOP will produce a number of arrivals resulting from multiple propagation paths for any source-receiver pair.
Using only the $N_0$ rays with neither surface nor bottom bounces, it will then estimate the current group velocity $u$ from a power weighted average of the ray travel times,
\begin{equation}
u = \frac{\hat{r} \sum_{n=1}^{N_{0}} a_{n}^{2}}{\sum_{n=1}^{N_{0}} dt_{n}a_{n}^{2}} ~, 
\end{equation}
and the associated weighted standard deviation,
\begin{equation}
\sigma_{u} \simeq \sqrt{\frac {\sum_{n=1}^{N_{0}} (dt_{n}-\hat{r}/u)^{2}a_{n}^{2}}{ \sum_{n=1}^{N_{0}} a_{n}^{2}} } \frac{u^{2}}{\hat{r}}
\end{equation}
If no direct paths exist, i.e. $N_{0}=0$, then the group velocity is computed using the same algorithm for the ray arrivals with one bounce, and so on.

@@

This stochastic method for group velocity calculation can run in real-time, appearing to be orders of magnitude faster than general post-processing methods which seek to determine the specific ray itself that best matches a prominent indicator from the arrival structure.
The method assumes direct path arrival, where multipath over the shallow receiver depth is trivial for the group velocity calculation; thus we call it the minimal bounce criteria (MBC).
The BELLHOP simulation that runs this calculation uses 3600 rays with launch angle fan of -60 to 60 degrees.
However, this high density of launch angles creates a greedy estimator of direct path eigenrays, and this simulation pipeline almost always guarantees a direct path found.
The MBC improves upon a deterministic estimate in that it considers the simplest ray path, travel time, and amplitude to predict a realistic group velocity.

@@

As shown in the eigenray traces of Fig. \ref{fig:raytrace-zs30}, an acoustic arrival does not always take the direct path from source to the receiver, and the path it takes is extremely dependent on the sound speed profile.
The suggested algorithm, the nearest bounce criteria (NBC), is a slight modification from the minimal bounce criteria, and includes multipath as a new dimension of information with negligible additional computation.
This metric was introduced in post-processing and does not include the plane wave approximation.

@@

This method selects a different group velocity based on the multipath arrival structure, as the expected arrival is not always the first arrival and could even be masked by noise or blocked temporarily \citep{deffenbaugh_acoustic_1996}.
We manually cap the number of bounces to 4 because of the smaller operational scale and the attenuation accrued with many surface interactions.
Bottom bounces are not encoded separately because of ray's tendency to refract upward, not due to information limitations. 
In addition, for a more consistent comparison, the local sparse grid is standardized as 11$x$11 points over 10 m in range and 20 m in depth, which is similar to those from the vehicle logs.
The extra size in depth addresses the computational limitations of how time-bounded eigenrays seem to stack vertically in the water column, as shown in Fig \ref{fig:rays3env-zs30}.
Lastly, to populate the multipath criterion effectively with a multitude of arrivals per number of surface bounces, we abandon the plane-wave extension to the interpolated receiver location.

@@

\FloatBarrier
\section{\label{sec:realtime} Real-time psuedorange results}

The modem experiment generated 811 beacon to beacon communication events with their own real-time NBC group velocity predictions.
As expected, the algorithm is generally overestimating range as it resolves a direct path that does not represent the actual arrival.
This section examines the absolute range error between the psuedorange and the GPS derived range with the caveat that the data collected is fairly unbalanced across source depths and sound speed conditions.

@@

The minimal and nearest bounce algorithms are applied, both without the plane wave extensions, with the three sound speed inputs shown in Fig.~\ref{fig:raytrace-zs30}.
The resulting predicted group velocities for a source depth of 30 m are shown in Fig~\ref{fig:gvel30}.
The takeaways from this source depth are representative of those from 20 and 90 m as well (see Supplementary Fig.~\ref{fig:gvelMore}).

@@

% =========================================================================== %
% =========================================================================== %
\section{\label{sec:discussion} Discussion}
